---
title: "Sample size calculation"
author: "Author"
date: "21 4 2022"
output:
  github_document: default
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE)
```


# Background

In clinical research, one of the first technical hurdles to a project is the sample size calculation - how many participants would you need to include to be able to answer your research question? They are required by funding bodies, ethics and safety regulators. They can also be important to the sponsor - without a good sample size calculation, how can you tell if you will be able to achieve your aims?

<!-- Figure here of clinician/statistician discussion?? Michael Coslovsky used to have a funny one... -->

Speaking with a statistician about sample size calculation can be frustrating for both parties. The investigator just wants a number that they can write in their proposal/protocol/grant application. The statistician cannot give a number without various details. Hence, a good understanding from both sides is primordial.
The aim of the study has to be well defined and all the needed assumptions educated guessed.

### Two main paradigms:

#### Null hypothesis significance testing (NHST) 

The aim is to assess if there is a statistical difference between e.g. two groups. For example

- does medication X lower Y more than medication Z
- does surgery X results in fewer Y than surgery Z

#### Precision based approaches
The focus is rather on estimating a quantity to a given accuracy. For example, we might expect that 40% of people have a given event within a certain time frame, but we want to estimate that 40% with a specific precision, in other words within a certain interval, e.g. 5% (2.5% on either side). We are interested in the width of the confidence interval (CI). It is also possible to estimate width of the CI for differences (e.g. mean differences) and relative effects (e.g. odds ratios, relative risks).

### What the statistician needs to know
#### What is the main comparison?

Between two groups is the most common (e.g. intervention vs control).

#### What is the primary endpoint? 
- what type of variable is it? Continuous? Binary? Time-to-event?
- when is it measured? Thirty days? Six months? One year?

#### What effect do you expect to see? 
- if you have continuous data, will the difference in means be 10, 20, ..., 100?
- if you have binary data, what proportion of events do you expect in the groups?
  
#### How variable will the endpoint be?
For continuous outcomes, standard deviations are typically called for. This is less relevant if you have a binary endpoint.

<!-- Panel A in the following figure show densities with identical mean values (10), but different standard deviations. Testing a difference with the one on the left would require a much smaller sample size than for the case on the right. Data from those two distributions are shown in panel B. -->


```{r, include = FALSE}
df <- data.frame(x = rep(seq(0,20, length.out = 200), times = 2),
                 g = factor(rep(c(1,2), each = 200), 1:2, c("Low SD", "High SD")))
df$d <- c(dnorm(df$x[1:200], 10, 1), dnorm(df$x[201:400], 10, 5))
df$r <- c(rnorm(df$x[1:200], 10, 1), rnorm(df$x[201:400], 10, 5))



library(ggplot2)

g1 <- ggplot(df, aes(x = x, y = d, group = g)) +
  geom_line() +
  facet_wrap(~ g)

bw <- .75
ggplot(df, aes(x = r)) +
  geom_histogram(col = "white") +
  geom_density(aes(y = ..density.. * (200 * bw)), col = "red") +
  facet_wrap(~ g)

g2 <- ggplot(df, aes(y = r, x = g)) +
  geom_point(position = position_jitter(height = 0))


library(patchwork)

g1 + g2 + plot_layout(2, width = c(.7, .3)) + plot_annotation(tag_levels = "A")

```




# Example(s)

Here we present a couple of very short examples of how sample size calculations can be performed, although we do recommend discussing your project with a statistician.

### The NHST framework
Suppose we want to estimate the mean difference between two groups (intervention vs placebo) and would like to be able to test it against "no difference". 
We have to define which difference will be clinically relevant to test. Let assume that a difference of 20 between both groups is clinically relevant. We also need to guess in an educated way the standard deviation (SD) of the outcome in the groups. Let assume a SD of 12 in each group.
We also have to set the power we want to achieve and the probability of type I error (alpha) we will allowed. In this example a power of 90% with a two-sided alpha of 5% can be reached with a sample size of 18, i.e. 9 participants in each group, to test if this difference is equal to null.

Parameters to assume or set:

- mean difference (delta)
- standard deviations
- power
- probability of type I error (alpha)

```{r}
power.t.test(delta = 20, sd = 12, sig.level = 0.05, power = .9,
             type = "two.sample",
             alternative = "two.sided")
```

If the difference between both groups is smaller the requested sample size to reach the same power with all the other parameters as above will be larger. Similarly, increasing the larger variation  around the mean (larger standard deviation) also increases the sample size.

```{r}
power.t.test(delta = 10, sd = 12, sig.level = 0.05, power = .9,
             type = "two.sample",
             alternative = "two.sided")
```

Multiple plausible scenario can be investigated, by varying some parameters. The [`sse` package](https://CRAN.R-project.org/package=sse) (@sse), provides a mechanism to create multiple scenarios and examine power in each, together with functions to aid reporting. `sse` is very flexible, but does require some programming. 

```{r, message = FALSE}
library(sse)

## defining the range of n and theta to be evaluated
psi <- powPar(
  # SD values
  theta = seq(from = 5, to = 20, by = 1),
  # sample sizes
  n = seq(from = 5, to = 50, by = 2),
  # group means
  muA = 0,
  muB = 20)

## define a function to return the power in each scenario
powFun <- function(psi){
  power.t.test(n = n(psi)/2,
               delta = pp(psi, "muA") - pp(psi, "muB"),
               sd = theta(psi)
  )$power
}

## evaluate the power-function for all combinations of n and theta
calc <- powCalc(psi, powFun)

## choose one particular example at theta of 1 and power of 0.9
pow <- powEx(calc, theta = 12, power = 0.9)

## drawing the power plot with 3 contour lines
plot(pow,
     xlab = "Standard Deviation",
     ylab = "Total Sample Size",
     at = c(0.85, 0.9, 0.95))
```

For additional details, see the `sse` package documentation on [CRAN](https://CRAN.R-project.org/package=sse).

### The precision based framework

Suppose we want to estimate the proportion of deaths within 30 days of a myocardial infarction (MI). @law2002 estimated that 36% (31% - 40%) of patients died within 30 days of an MI. If we want to replicate their study, we can estimate the number of participants that would be necessary to achieve a CI which is 9% wide. Members of the SCTO Statistics and Methodology platform created an R package specifically for this problem called `presize` (@presize). It is available on [CRAN](https://CRAN.R-project.org/package=presize) or as an easy to use [shiny application](https://shiny.ctu.unibe.ch/presize/).

The relevant function in presize is the `prec_prop` function. Plugging in the numbers from above, we can see that we require 434 participants (column `n`) to yield a confidence interval from 31.6% (`lwr`) to 40.6% (`upr`).

```{r}
library(presize)
prec_prop(p = 0.36, conf.width = 0.09)
```

If we know that we can only afford to follow 300 participants, we can see what the confidence interval would be in that case too, via the `n` argument (instead of `cond.width`):

```{r}
prec_prop(p = 0.36, n = 300)
```

It's also possible to check multiple scenarios, by passing multiple values to a parameter. Below we vary the number of participants (what CI width is possible to determine with that number of participants?):
```{r}
prec_prop(p = .36, n = c(300, 350, 400, 450, 450))
```

The results can also be plotted easily, which is particularly useful when running multiple scenarios.

```{r}
library(ggplot2)  # for plotting
library(magrittr) # for 'piping'

prec_prop(p = .36, n = c(300, 350, 400, 450, 500)) %>% 
  as.data.frame() %>% 
  ggplot(aes(x = n, y = conf.width)) +
  geom_line() +
  labs(x = "N participants", 
       y = "CI width")
```

For additional details and examples, see the [`presize` website](https://ctu-bern.github.io/presize/).

# Conclusion

Sample size calculation is a very important part of the study planning process. To make the most of your meeting with the statistician, have the answers to the questions mentioned above.

# References
